What to simulate? Inferring the right direction for mental rotation

### 2014-02-05 man/cogsci-2014/cogsci-2014.tex
Response time --> RT

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/response_time_accuracy.pdf}
    \caption{\textbf{Response time and accuracy comparison.} Top:
      response time of correct responses as a function of the minimum
      angle of rotation. Bottom: accuracy as a function of the minimum
      angle of rotation. All error bars are 95\% confidence
      intervals.}
    \label{fig:response-time-accuracy}
  \end{center}
\end{figure*}

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/response_time_scatters.pdf}
    \caption{\textbf{Model vs. human response times.}}
    \label{fig:human-model-scatters}
  \end{center}
\end{figure*}

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the rest, we excluded 10 (4\%) because of an experimental
error, 6 (2.4\%) because they had already completed a related
experiment, and 31 (12.6\%) because they failed a comprehension
check. Comprehension was defined as correctly answering at least 85\%
of stimuli with a minimum rotation of $0^\circ$ or $20^\circ$.
Additionally, we excluded 82 trials for which the response time was
either less than 100 milliseconds or greater than 20 seconds.

For each model, we ran 50 samples for each of the 800 experimental
stimuli. The step size parameter ($\sigma_\epsilon$) was fit to human
response times for each of the models, resulting in
$\sigma_\epsilon=0.6$ for the \Th{} and \Bq{} models and
$\sigma_\epsilon=0.1$ for the \Oc{} and \Hc{} models. We also ran the
models under two different priors, $p(h=0)=0.5$ (the ``equal'' prior)
and $p(h=0)=0.55$ (the ``unequal'' prior). This only had an effect on
the stopping criteria for the \Bq{} model.\footnote{The prior also had
  a small effect on the accuracy of the \Hc{} model, because sometimes
  the local maxima were small enough that Eq.~\ref{eq:mle-lh-ratio}
  became dominated by the prior.}

@@

For analyses of response time, confidence intervals around harmonic
means were computed using a bootstrap analysis using 10000 bootstrap
samples (sampled with replacement). Response times were only
considered for correct responses.  We also used a bootstrap analysis
of 10000 bootstrap samples to compute the confidence intervals around
both Spearman ($r_s$) and Pearson ($\rho$) correlations.  Unless
otherwise specified, all correlations were computed over 720
datapoints (one for each stimulus). For analyses of accuracy,
confidence intervals were computed from a binomial proportion with a
Jeffrey's beta prior.  To test if participants or a model was above
chance on a particular stimulus, we used the same binomial proportion,
and simply tested whether $p>0.05$ for the 50\% mark.

@@

The average response time across all stimuli was \ExpTime{}
milliseconds.  The minimum angle of rotation was significantly
monotonically correlated with average per-stimulus response times,
both for ``flipped'' and ``same'' pairs \TODO{}. While this replicates
the general result of previous experiments
\cite<e.g.,>{Shepard1971,Cooper:1975wp}, our results are not as
perfectly linear. However, \citeA{Gardony:2013gn} found a constant
effect for ``flipped'' pairs; an average of their results and the
classic results would produce something similar to ours.

@@

To gauge between-subject reliability, we randomly split subjects into
two groups, computed the average response time and accuracy per
stimulus for each group, and then computed correlations for these two
measures. We repeated this procedure 100 times, and from those samples
computed the median correlations and 95\% confidence intervals. For
response time, the between-subjects correlation was \ExpTimeCorr{},
and for accuracy, it was \ExpAccuracyCorr{}.

There was a significant effect of trial number both on response time
(\ExpTrialTimeCorr{}) and on accuracy (\ExpTrialAccuracyCorr{}),
though the effect on accuracy was not significant during the second
half of the experiment (\ExpaTrialAccuracyCorr{} for the first half
vs. \ExpbTrialAccuracyCorr{} for the second half). These practice
effects may have contributed to the not-quite-linearity of the human
response times; future work will need to collect more data per
participant to mitigate this effect.

@@

As we would expect, the number of actions taken by the \Oc{} model was
perfectly correlated with the minimum angle of rotation
(Fig.~\ref{fig:response-time-accuracy}). Unsurprisingly, the \Oc{}
model was the best fit to human response times, with a correlation of
\OcTimeCorr{} (Fig.~\ref{fig:human-model-scatters}).  However, the
\Oc{} model was 100\% accurate, and therefore the correlation between
minimum angle and accuracy was undefined.

@@

There was an overall monotonic relationship between the minimum angle
of rotation and the number of actions taken by the \Th{} model
(Fig.~\ref{fig:response-time-accuracy}). However, the relationship
between angle and response time for individual shapes was usually not
monotonic (e.g., Fig.~\ref{fig:response-time-stimulus}).  The \Th{}
model was able to explain a moderate amount of the variance in human
response times, with a correlation of \ThTimeCorr{}
(Fig.~\ref{fig:human-model-scatters}). Like the \Oc{} model, the \Th{}
model had 100\% accuracy, and thus there was no correlation between
minimum angle and accuracy.

@@

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{../../figures/D/response_time_stimulus.pdf}
    \caption{\textbf{Typical response time curves for a single shape.}
      These plots correspond to the shape shown in
      Fig.~\ref{fig:model-traces}. Left: human curves are either
      linear (as with the ``same'' pairs), or linear and then flat (as
      with the ``flipped'' pairs). Middle: the \Th{} model tends to
      have highly inconsitent within-shape response times. Right: the
      \Bq{} model is roughly linear.}
    \label{fig:response-time-stimulus}
  \end{center}
\end{figure}

The \Hc{} was the only model for which there was no monotonic
relationship between rotation and response time
(Fig.~\ref{fig:response-time-accuracy}). Moreover, the \Hc{} model was
barely above chance (\HcAccuracy{}) and there were \HcNumChance{}
individual stimuli for which it was not above chance. The \Hc{} model
was not a particularly good predictor of human response times
(\HcTimeCorr{}), as shown in Fig.~\ref{fig:human-model-scatters}. It
was a moderately good predictor of human accuracy (\HcAccuracyCorr{}).

@@

Like the \Oc{} and \Th{} models, there was an overall monotonic
relationship between rotation and the number of steps taken by the
\Bq{} model (Fig.~\ref{fig:response-time-accuracy}). Unlike the \Th{}
model, this relationship existed for individual shapes as well (e.g.,
Fig.~\ref{fig:response-time-stimulus}).  The \Bq{} model explained
variance in human response times about as well as the \Th{} model
(Fig.~\ref{fig:human-model-scatters}), with a correlation of
\BqTimeCorr{} for the equal prior and \BqpTimeCorr{} for the unequal
prior.

@@

Because the \Bq{} model relies on Eq.~\ref{eq:lh-ratio} for its
stopping criteria (as opposed to just finding a maximum), the prior
$p(h)$ has a fairly significant effect, as shown in
Fig.~\ref{fig:response-time-accuracy}. With just a small bias of
$p(h=0)=0.55$, there is a clear separation in response times for
``same'' versus ``flipped'' stimuli; this separation is similar to the
trend also observed in human response times. While the prior also had
an effect on accuracy, this effect did not reflect the trends in human
behavior.

@@

Additionally, a closer look at performance on individual shapes
reveals several interesting points
(Fig.~\ref{fig:response-time-stimulus} shows curves for one particular
shape; these curves are fairly typical relative to other
shapes). First, the curves for the human response times on specific
shapes tend to either be linear, or linear first, then flat. Second,
while the \Th{} model appears to capture the concavity of human
response times overall, it is actually fairly inconsistent within
shapes. Third, the linearity of the \Bq{} model's response time curves
\textit{is} consistent within shapes, though it does not exhibit the
same ``plateau'' effect that sometimes appears in the human curves.

>>>>

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/response_time_accuracy.pdf}
    \caption{\textbf{Response time and accuracy comparison.} Top: RT
      of correct responses as a function of the minimum angle of
      rotation. Bottom: accuracy as a function of the minimum angle of
      rotation. All error bars are 95\% confidence intervals.}
    \label{fig:response-time-accuracy}
  \end{center}
\end{figure*}

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/response_time_scatters.pdf}
    \caption{\textbf{Model vs. human RTs.}}
    \label{fig:human-model-scatters}
  \end{center}
\end{figure*}

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the rest, we excluded 10 (4\%) because of an experimental
error, 6 (2.4\%) because they had already completed a related
experiment, and 31 (12.6\%) because they failed a comprehension
check. Comprehension was defined as correctly answering at least 85\%
of stimuli with a minimum rotation of $0^\circ$ or $20^\circ$.
Additionally, we excluded 82 trials for which the response time (RT)
was either less than 100 milliseconds or greater than 20 seconds.

For each model, we ran 50 samples for each of the 800 experimental
stimuli. The step size parameter ($\sigma_\epsilon$) was fit to human
RTs for each of the models, resulting in $\sigma_\epsilon=0.6$ for the
\Th{} and \Bq{} models and $\sigma_\epsilon=0.1$ for the \Oc{} and
\Hc{} models. We also ran the models under two different priors,
$p(h=0)=0.5$ (the ``equal'' prior) and $p(h=0)=0.55$ (the ``unequal''
prior). This only had an effect on the stopping criteria for the \Bq{}
model.\footnote{The prior also had a small effect on the accuracy of
  the \Hc{} model, because sometimes the local maxima were small
  enough that Eq.~\ref{eq:mle-lh-ratio} became dominated by the
  prior.}

@@

For analyses of RT, confidence intervals around harmonic means were
computed using a bootstrap analysis using 10000 bootstrap samples
(sampled with replacement). RTs were only considered for correct
responses.  We also used a bootstrap analysis of 10000 bootstrap
samples to compute the confidence intervals around both Spearman
($r_s$) and Pearson ($\rho$) correlations.  Unless otherwise
specified, all correlations were computed over 720 datapoints (one for
each stimulus). For analyses of accuracy, confidence intervals were
computed from a binomial proportion with a Jeffrey's beta prior.  To
test if participants or a model was above chance on a particular
stimulus, we used the same binomial proportion, and simply tested
whether $p>0.05$ for the 50\% mark.

@@

The average RT across all stimuli was \ExpTime{} milliseconds.  The
minimum angle of rotation was significantly monotonically correlated
with average per-stimulus RTs, both for ``flipped'' and ``same'' pairs
\TODO{}. While this replicates the general result of previous
experiments \cite<e.g.,>{Shepard1971,Cooper:1975wp}, our results are
not as perfectly linear. However, \citeA{Gardony:2013gn} found a
constant effect for ``flipped'' pairs; an average of their results and
the classic results would produce something similar to ours.

@@

To gauge between-subject reliability, we randomly split subjects into
two groups, computed the average RT and accuracy per stimulus for each
group, and then computed correlations for these two measures. We
repeated this procedure 100 times, and from those samples computed the
median correlations and 95\% confidence intervals. For RT, the
between-subjects correlation was \ExpTimeCorr{}, and for accuracy, it
was \ExpAccuracyCorr{}.

There was a significant effect of trial number both on RT
(\ExpTrialTimeCorr{}) and on accuracy (\ExpTrialAccuracyCorr{}),
though the effect on accuracy was not significant during the second
half of the experiment (\ExpaTrialAccuracyCorr{} for the first half
vs. \ExpbTrialAccuracyCorr{} for the second half). These practice
effects may have contributed to the not-quite-linearity of the human
RTs; future work will need to collect more data per participant to
mitigate this effect.

@@

As we would expect, the number of actions taken by the \Oc{} model was
perfectly correlated with the minimum angle of rotation
(Fig.~\ref{fig:response-time-accuracy}). Unsurprisingly, the \Oc{}
model was the best fit to human RTs, with a correlation of
\OcTimeCorr{} (Fig.~\ref{fig:human-model-scatters}).  However, the
\Oc{} model was 100\% accurate, and therefore the correlation between
minimum angle and accuracy was undefined.

@@

There was an overall monotonic relationship between the minimum angle
of rotation and the number of actions taken by the \Th{} model
(Fig.~\ref{fig:response-time-accuracy}). However, the relationship
between angle and RT for individual shapes was usually not monotonic
(e.g., Fig.~\ref{fig:response-time-stimulus}).  The \Th{} model was
able to explain a moderate amount of the variance in human RTs, with a
correlation of \ThTimeCorr{}
(Fig.~\ref{fig:human-model-scatters}). Like the \Oc{} model, the \Th{}
model had 100\% accuracy, and thus there was no correlation between
minimum angle and accuracy.

@@

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{../../figures/D/response_time_stimulus.pdf}
    \caption{\textbf{Typical RT curves for a single shape.}  These
      plots correspond to the shape shown in
      Fig.~\ref{fig:model-traces}. Left: human curves are either
      linear (as with the ``same'' pairs), or linear and then flat (as
      with the ``flipped'' pairs). Middle: the \Th{} model tends to
      have highly inconsitent within-shape RTs. Right: the \Bq{} model
      is roughly linear.}
    \label{fig:response-time-stimulus}
  \end{center}
\end{figure}

The \Hc{} was the only model for which there was no monotonic
relationship between rotation and RT
(Fig.~\ref{fig:response-time-accuracy}). Moreover, the \Hc{} model was
barely above chance (\HcAccuracy{}) and there were \HcNumChance{}
individual stimuli for which it was not above chance. The \Hc{} model
was not a particularly good predictor of human RTs (\HcTimeCorr{}), as
shown in Fig.~\ref{fig:human-model-scatters}. It was a moderately good
predictor of human accuracy (\HcAccuracyCorr{}).

@@

Like the \Oc{} and \Th{} models, there was an overall monotonic
relationship between rotation and the number of steps taken by the
\Bq{} model (Fig.~\ref{fig:response-time-accuracy}). Unlike the \Th{}
model, this relationship existed for individual shapes as well (e.g.,
Fig.~\ref{fig:response-time-stimulus}).  The \Bq{} model explained
variance in human RTs about as well as the \Th{} model
(Fig.~\ref{fig:human-model-scatters}), with a correlation of
\BqTimeCorr{} for the equal prior and \BqpTimeCorr{} for the unequal
prior.

@@

Because the \Bq{} model relies on Eq.~\ref{eq:lh-ratio} for its
stopping criteria (as opposed to just finding a maximum), the prior
$p(h)$ has a fairly significant effect, as shown in
Fig.~\ref{fig:response-time-accuracy}. With just a small bias of
$p(h=0)=0.55$, there is a clear separation in RTs for ``same'' versus
``flipped'' stimuli; this separation is similar to the trend also
observed in human RTs. While the prior also had an effect on accuracy,
this effect did not reflect the trends in human behavior.

@@

Additionally, a closer look at performance on individual shapes
reveals several interesting points
(Fig.~\ref{fig:response-time-stimulus} shows curves for one particular
shape; these curves are fairly typical relative to other
shapes). First, the curves for the human RTs on specific shapes tend
to either be linear, or linear first, then flat. Second, while the
\Th{} model appears to capture the concavity of human RTs overall, it
is actually fairly inconsistent within shapes. Third, the linearity of
the \Bq{} model's RT curves \textit{is} consistent within shapes,
though it does not exhibit the same ``plateau'' effect that sometimes
appears in the human curves.
